---
title: "Technical Appendix"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: show
    theme: cosmo
---

```{r}
library(tidycensus) 
library(tidyverse)
library(here)
library(lubridate)
library(dplyr)
library(sf)
library(tigris)
library(spatstat)
library(terra)
library(ggplot2)
library(arrow)
library(tigris)
library(car)
library(lmtest)
library(spdep)
library(ggpubr)
library(ggplot2)
library(cowplot)
library(MASS)
```
```{r}
df=read.csv(here("data/philadelphia_monthly_2020_2021.csv"))
df$week_date <- my(df$month)
df$first_of_month=floor_date(df$week_date,"month")
summary(df)
```


```{r}
acs_tables <- load_variables(year = 2023, dataset = "acs5", cache = TRUE)
```

```{r}
census_path <- here("data", "Philly Census")

census_shp_path <- file.path(census_path, "philly_tract.shp")

philly_tract_sf <- st_read(census_shp_path, quiet = TRUE)
```


Variables Collected:

Median gross rent: (B25064)

Race: (B02001)

Median household income: (B19013)

% of Renter: (B25003)

Year Built(renter): (B25036)

Number of Room: (B25017)

Room Per Household: (B25020)

Family Structure: (B11001)

Unemployment in labor force: (C18120)

Gross Rent as a Percentage of Household Income: (B25070)

### census data collected

```{r api key, eval=TRUE, include=FALSE}

census_api_key <- ("138db656646183a18ae51e4a7c1e1f5cd64d4b40")

# Define parameters
my_state  <- "Pennsylvania"
my_county <- "Philadelphia"
acs_year  <- 2023

# Median Gross Rent
median_rent <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  variables = "B25064_001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, median_rent = estimate)

# Race
race <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  table = "B02001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    total_pop = B02001_001,
    white = B02001_002,
    black = B02001_003,
    asian = B02001_005,
    pct_white = 100 * white / total_pop,
    pct_black = 100 * black / total_pop,
    pct_asian = 100 * asian / total_pop
  ) %>%
  select(GEOID, total_pop, pct_white, pct_black, pct_asian)

# Median Household Income
median_income <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  variables = "B19013_001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, median_hh_income = estimate)

# Percent Renter
tenure <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  table = "B25003",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    total_occupied = B25003_001,
    renter_occupied = B25003_003,
    pct_renter = 100 * renter_occupied / total_occupied
  ) %>%
  select(GEOID, pct_renter)

# Median Year Built - Renter Occupied
year_built_renter <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  variables = "B25037_003",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, median_year_built_renter = estimate)

# Number of Rooms
num_rooms <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  variables = "B25017_001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, aggregate_rooms = estimate)

# Median Rooms Per Household 
rooms_per_hh <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  variables = "B25018_001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, median_rooms = estimate)

# Family Structure 
family_structure <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  table = "B11001",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    total_hh = B11001_001,
    family_hh = B11001_002,
    nonfamily_hh = B11001_007,
    pct_family = 100 * family_hh / total_hh
  ) %>%
  select(GEOID, total_hh, pct_family)

# Unemployment in Labor Force (C18120)
unemployment <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  table = "C18120",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    in_labor_force = C18120_002,
    unemployed = C18120_004,
    pct_unemployed = 100 * unemployed / in_labor_force
  ) %>%
  select(GEOID, pct_unemployed)

# Gross Rent as Percentage of Household Income ()
rent_burden <- get_acs(
  geography = "tract",
  state = my_state,
  county = my_county,
  table = "B25070",
  year = acs_year,
  survey = "acs5"
) %>%
  select(GEOID, variable, estimate) %>%
  tidyr::pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    total = B25070_001,
    # Rent burdened = paying 30%+ of income on rent
    rent_burdened = B25070_007 + B25070_008 + B25070_009 + B25070_010,
    pct_rent_burdened = 100 * rent_burdened / total
  ) %>%
  select(GEOID, pct_rent_burdened)

#Join All data
philly_tract_data <- median_rent %>%
  left_join(race, by = "GEOID") %>%
  left_join(median_income, by = "GEOID") %>%
  left_join(tenure, by = "GEOID") %>%
  left_join(year_built_renter, by = "GEOID") %>%
  left_join(num_rooms, by = "GEOID") %>%
  left_join(rooms_per_hh, by = "GEOID") %>%
  left_join(family_structure, by = "GEOID") %>%
  left_join(unemployment, by = "GEOID") %>%
  left_join(rent_burden, by = "GEOID")

philly_tract=tracts( state = 'PA',
        county = '101',
        year = 2023
)

philly_tract_sf=philly_tract%>%
  select(GEOID,geometry)
```


### Dewey rental data
```{r}
#input data
folder_path <- "data/rental-in-philly"
all_parquet_files <- open_dataset(
  sources = folder_path, 
  format = "parquet"
)
combined_df <- all_parquet_files %>%
  collect()

#reproject it
rental_shp=combined_df%>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

#input city limit data  
api_url <- "https://services.arcgis.com/fLeGjb7u4uXqeF9q/arcgis/rest/services/City_Limits/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
city_limits_data <- st_read(api_url)


rental_shp_proj <- st_transform(rental_shp, crs = 32618)
city_limits_data_proj <- st_transform(city_limits_data, crs = 32618)

#filter to philadelphia
phillyrental_proj <- st_filter(rental_shp_proj, city_limits_data_proj, .pred = st_within)

#filter to different times
phillyrental_proj_1923=phillyrental_proj%>%
  filter(
  year(DATE_POSTED) %in% c(2019:2023)
)

phillyrental_proj_1619=phillyrental_proj%>%
  filter(
  year(DATE_POSTED) %in% c(2016:2019)
)

philly_tract_sf=st_transform(philly_tract_sf,crs=32618)

#merge with census tract
phillyrental_proj_1923=phillyrental_proj_1923%>%
  st_join(philly_tract_sf,join = st_within)

phillyrental_proj_1619=phillyrental_proj_1619%>%
  st_join(philly_tract_sf,join = st_within)


```


### load quartly wage data
```{r}
file_list <- list.files(
  path = "data",
  pattern = ".q1-q4 42101 Philadelphia County, Pennsylvania.csv",
  full.names = TRUE
)
wage_data <- file_list %>%
  set_names() %>% 
  map_df(~ read_csv(.x), .id = "source_file")%>%
  filter(own_title=="Total Covered")%>%
  select(year,qtr,avg_wkly_wage)
```

### merge wage data with evictation data
```{r}
df$quarter=quarter(df$week_date)
df$year=year(df$week_date)
df$yearqu=paste0(df$year, df$quarter)
wage_data$yearqu=paste0(wage_data$year,wage_data$qtr)
df=wage_data%>%
  select(yearqu,avg_wkly_wage)%>%
  right_join(df,by="yearqu")
  
```

### merge with different CPI and unemployment rate per month in Philadelphia
```{r}
food=read.csv(here('data/food.csv'))
electricity=read.csv(here('data/electricity.csv'))
commodities=read.csv(here('data/commodities.csv'))

#merge with electricity cpi
df$month=month(df$week_date)
electricity$month=as.numeric(gsub("M","",electricity$Period))
df$yearm=paste0(df$year,df$month)
electricity$yearm=paste0(electricity$Year,electricity$month)

df=electricity%>%
  mutate(electricitycpi=Value)%>%
  select(yearm,electricitycpi)%>%
  right_join(df,by="yearm")

#merge with food cpi
food$first_of_month=ym(food$Label)
df=food%>%
  mutate(foodcpi=Value)%>%
  select(first_of_month,foodcpi)%>%
  right_join(df, by="first_of_month")

df=df%>%
  arrange(week_date)%>%
  fill(foodcpi, .direction = "up")

#merge with commodities cpi

commodities = commodities %>%
  mutate(
    comcpi = Value, 
    half_year_date = case_when(
      Period == "S01" ~ ymd(paste(Year, "01", "01")), # Half 1 -> Jan 1
      Period == "S02" ~ ymd(paste(Year, "07", "01"))  # Half 2 -> Jul 1
    )
  ) %>%
  select(half_year_date, comcpi)

df = df %>%
  left_join(commodities, by = c("first_of_month" = "half_year_date"))

df = df %>%
  arrange(week_date) %>%
  fill(comcpi, .direction = "down")

```

### merge with unemployment data
```{r}
unemployment=read.csv("data/unemployment.csv")
unemployment$week_date=ym(unemployment$Label)
df=unemployment%>%
  mutate(unemployment_rate=Value)%>%
  select(unemployment_rate,week_date)%>%
  right_join(df,by="week_date")
```


### merge with census data 
```{r}
df_census=df%>%
  left_join(philly_tract_data,by="GEOID")


df_census_sf=philly_tract_sf%>%
  left_join(df_census,by="GEOID")
```

### merge with neighborhood data
```{r}
neighborhood_folder <- here("data", "philadelphia-neighborhoods")
neighborhood_path   <- file.path(neighborhood_folder, "philadelphia-neighborhoods.shp")




# Read the shapefile
philly_neighborhoods <- st_read(neighborhood_path, quiet = TRUE)

philly_neighborhoods=st_transform(philly_neighborhoods, crs=32618)
df_census_sf=st_join(df_census_sf,philly_neighborhoods)
census_points=st_centroid(df_census_sf)

points_joined <- st_join(census_points, philly_neighborhoods)

# Calculate mean MHI for each polygon
mean_mhi_by_poly <- points_joined %>%
  st_drop_geometry() %>%             
  group_by(MAPNAME) %>%            
  summarise(meanMHI = mean(median_hh_income, na.rm = TRUE))

# Join the result back to the polygon layer
df_census_sf_final <- df_census_sf %>%
  left_join(mean_mhi_by_poly, by = "MAPNAME")%>%
  st_as_sf() 
# reclassify the neighborhood data based on quantile (25%)
df_census_sf_final$MHI_quantile <- cut(
  df_census_sf_final$meanMHI,
  breaks = quantile(
    df_census_sf_final$meanMHI,
    probs = seq(0, 1, 0.25),   
    na.rm = TRUE
  ),
  include.lowest = TRUE,
  labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)")
)

```




### merge with rent data
```{r}
#calculate rent by month and tract
rent_by_month_tract=phillyrental_proj_1923%>%
  st_drop_geometry()%>%
  mutate(date=format(DATE_POSTED,"%Y-%m"))%>%
  select(GEOID,RENT_PRICE,date,NEIGHBORHOOD)%>%
  group_by(GEOID,date)%>%
    summarise(median_rent=median(RENT_PRICE))
```


### Final data frame

```{r}
#merge with eviction data
df_final=df_census_sf_final%>%
  mutate(date=format(week_date,"%Y-%m"))%>%
  filter(year(week_date) %in% c(2020:2023))%>%
  left_join(rent_by_month_tract,by=c("date","GEOID"))%>%
  select(GEOID,unemployment_rate,foodcpi,electricitycpi,avg_wkly_wage,
         filings_2020,comcpi,total_pop,pct_white,pct_black,pct_asian,
         median_hh_income,pct_renter,median_year_built_renter,aggregate_rooms,
         median_rooms,total_hh,pct_family,median_rent.y,MHI_quantile,week_date,year)
```


#EXPLORATORY DATA ANALYSIS

```{r}
colnames(df_final)
```


# Monthly Eviction Filings by Year
```{r}
# Aggregate by year and month
monthly_by_year <- df %>%
  group_by(year, month) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE), .groups = "drop") %>%
  filter(year %in% c(2020, 2021, 2022, 2023))

# Check the data
print(monthly_by_year, n = 48)

# Function to create plot for each year
make_year_plot <- function(data, year_num) {
  year_data <- data %>% filter(year == year_num)
  
  ggplot(year_data) +
    geom_col(aes(x = factor(month), y = total_filings), 
             fill = "darkseagreen3", color = "black", width = 0.7) +
    scale_x_discrete(labels = month.abb) +
    labs(title = as.character(year_num),
         x = NULL,
         y = NULL) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
      axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 8)
    )
}

# Create 4 plots
plot_list <- map(c(2020, 2021, 2022, 2023), ~make_year_plot(monthly_by_year, .x))

# Arrange 2x2
combined <- ggarrange(plotlist = plot_list, ncol = 2, nrow = 2, align = 'hv')

annotate_figure(combined,
                top = text_grob("Monthly Eviction Filings by Year", 
                                face = "bold", size = 16),
                left = text_grob("Total Filings", rot = 90, size = 12))
```


# Eviction Filings by Month Across Years
```{r}
# aggregate by year and month
monthly_by_year <- df %>%
  group_by(year, month) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE), .groups = "drop") %>%
  filter(year %in% c(2020, 2021, 2022, 2023))

# Function to create plot for each month
make_month_plot <- function(data, month_num) {
  month_data <- data %>% filter(month == month_num)
  month_label <- month.abb[month_num]
  
  years_present <- sort(unique(month_data$year))
  year_labels <- substr(as.character(years_present), 3, 4)
  
  ggplot(month_data) +
    geom_col(aes(x = factor(year), y = total_filings), 
             fill = "darkseagreen3", color = "black", width = 0.7) +
    scale_x_discrete(labels = year_labels) +
    labs(title = month_label, x = NULL, y = NULL) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8)
    )
}

# Create 12 plots
plot_list <- map(1:12, ~make_month_plot(monthly_by_year, .x))

# Arrange 3x4
combined <- ggarrange(plotlist = plot_list, ncol = 4, nrow = 3, align = 'hv')

annotate_figure(combined,
                top = text_grob("Eviction Filings by Month Across Years", 
                                face = "bold", size = 14),
                left = text_grob("Total Filings", rot = 90, size = 11))
```


# Eviction Map
```{r}
# aggregate filings by GEOID and year
evictions_by_tract_year <- df %>%
  group_by(GEOID, year) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE), .groups = "drop") %>%
  filter(year %in% c(2020, 2021, 2022, 2023))

# join with census tract geometry
evictions_map_data <- philly_tract_sf %>%
  left_join(evictions_by_tract_year, by = "GEOID")

# function to create map for each year
make_year_map <- function(data, year_num) {
  year_data <- data %>% filter(year == year_num)
  
  ggplot(year_data) +
    geom_sf(aes(fill = total_filings), color = NA, lwd = 0) +
    scale_fill_viridis_c(
      option = "inferno",
      name = "Filings",
      na.value = "grey90"
    ) +
    labs(title = as.character(year_num)) +
    theme_void() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
      legend.position = "right",
      legend.key.height = unit(0.8, "cm"),
      legend.key.width = unit(0.3, "cm")
    )
}

# create 4 maps
map_list <- map(c(2020, 2021, 2022, 2023), ~make_year_map(evictions_map_data, .x))

combined_maps <- ggarrange(plotlist = map_list, ncol = 2, nrow = 2, 
                            common.legend = TRUE, legend = "right")

annotate_figure(combined_maps,
                top = text_grob("Eviction Filings by Census Tract (2020-2023)", 
                                face = "bold", size = 16))
```


# Local Moran's I
```{r}
# aggregate total filings by tract (across all years or pick one year)
evictions_by_tract <- df %>%
  filter(year == 2023) %>%
  group_by(GEOID) %>%
  summarise(total_filings = sum(filings_2020, na.rm = TRUE), .groups = "drop")
# join to geometry
evictions_sf <- philly_tract_sf %>%
  left_join(evictions_by_tract, by = "GEOID") %>%
  filter(!is.na(total_filings))
# create spatial weights (queen contiguity)
neighbors <- poly2nb(evictions_sf, queen = TRUE)
weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
# calculate Local Moran's I
local_moran <- localmoran(evictions_sf$total_filings, weights, zero.policy = TRUE)
# add results to sf object
evictions_sf <- evictions_sf %>%
  mutate(
    local_i = local_moran[, 1],
    p_value = local_moran[, 5],
    scaled_filings = scale(total_filings),
    lag_scaled = lag.listw(weights, scaled_filings, zero.policy = TRUE),
    # classify hot/cold spots
    hotspot_type = case_when(
      scaled_filings > 0 & lag_scaled > 0 & p_value < 0.05 ~ "Hot Spot (High-High)",
      scaled_filings < 0 & lag_scaled < 0 & p_value < 0.05 ~ "Cold Spot (Low-Low)",
      scaled_filings > 0 & lag_scaled < 0 & p_value < 0.05 ~ "High-Low Outlier",
      scaled_filings < 0 & lag_scaled > 0 & p_value < 0.05 ~ "Low-High Outlier",
      TRUE ~ "Not Significant"
    )
  )
# plot LISA map
ggplot(evictions_sf) +
  geom_sf(aes(fill = hotspot_type), color = NA) +
  scale_fill_manual(
    values = c(
      "Hot Spot (High-High)" = "#d7191c",
      "Cold Spot (Low-Low)" = "#2c7bb6",
      "High-Low Outlier" = "#fdae61",
      "Low-High Outlier" = "#abd9e9",
      "Not Significant" = "grey90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Eviction Hot Spots and Cold Spots (2023)",
    subtitle = "Local Moran's I Cluster Analysis",
    caption = "Hot spots = High evictions surrounded by high evictions"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  )
```


# Distance from hot spot
```{r}
#identify hot spot centroids from LISA analysis above
hotspot_tracts <- evictions_sf %>%
  filter(hotspot_type == "Hot Spot (High-High)")

hotspot_centroids <- st_centroid(hotspot_tracts)

# calculate distance from each tract to nearest hot spot
all_tracts <- philly_tract_sf %>%
  left_join(evictions_by_tract, by = "GEOID")

tract_centroids <- st_centroid(all_tracts)

# distance to nearest hot spot
dist_to_hotspot <- st_distance(tract_centroids, hotspot_centroids)
all_tracts$dist_to_hotspot <- apply(dist_to_hotspot, 1, min)
all_tracts$dist_to_hotspot_km <- as.numeric(all_tracts$dist_to_hotspot) / 1000

# plot distance map
ggplot(all_tracts) +
  geom_sf(aes(fill = dist_to_hotspot_km), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Distance\n(km)",
    direction = -1
  ) +
  geom_sf(data = hotspot_tracts, fill = NA, color = "red", size = 0.8) +
  labs(
    title = "Distance to Eviction Hot Spots",
    subtitle = "Red outlines = Hot spot tracts",
    caption = "Darker = closer to high-eviction clusters"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16))
```


# Eviction vs Unemployment Rate
```{r}
# prepare data
map_data <- df_final %>%
  group_by(GEOID) %>%
  summarise(
    total_filings = sum(filings_2020, na.rm = TRUE),
    unemployment = mean(unemployment_rate, na.rm = TRUE),
    .groups = "drop"
  )

map_sf <- philly_tract_sf %>%
  left_join(map_data, by = "GEOID") %>%
  filter(!is.na(total_filings) & !is.na(unemployment))

# create tertiles for each variable
map_sf <- map_sf %>%
  mutate(
    filings_class = ntile(total_filings, 3),
    unemp_class = ntile(unemployment, 3),
    bi_class = paste0(unemp_class, "-", filings_class)
  )

# define bivariate color palette
bivariate_colors <- c(
  "1-1" = "#e8e8e8",  # Low unemployment, Low evictions
  "2-1" = "#ace4e4",  # Med unemployment, Low evictions
  "3-1" = "#5ac8c8",  # High unemployment, Low evictions
  "1-2" = "#dfb0d6",  # Low unemployment, Med evictions
  "2-2" = "#a5add3",  # Med unemployment, Med evictions
  "3-2" = "#5698b9",  # High unemployment, Med evictions
  "1-3" = "#be64ac",  # Low unemployment, High evictions
  "2-3" = "#8c62aa",  # Med unemployment, High evictions
  "3-3" = "#3b4994"   # High unemployment, High evictions
)

# map
map1 <- ggplot(map_sf) +
  geom_sf(aes(fill = bi_class), color = NA) +
  scale_fill_manual(values = bivariate_colors, na.value = "grey90", guide = "none") +
  labs(
    title = "Eviction Filings vs Unemployment Rate",
    subtitle = "Darker purple = High evictions AND high unemployment"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16))

legend1 <- ggplot(legend_data, aes(x = x, y = y, fill = bi_class)) +
  geom_tile() +
  scale_fill_manual(values = bivariate_colors, guide = "none") +
  labs(x = "Unemployment →", y = "Evictions →") +
  theme_void() +
  theme(
    axis.title.x = element_text(size = 8, margin = margin(t = 8)),
    axis.title.y = element_text(size = 8, angle = 90, margin = margin(r = 8))
  ) +
  coord_fixed()

final_plot1 <- ggdraw() +
  draw_plot(map1, 0, 0, 1, 1) +
  draw_plot(legend1, 0.68, 0.02, 0.28, 0.28)

print(final_plot1)
```

# Eviction vs Median Gross Rent
```{r}
# prepare data
map_data <- df_final %>%
  group_by(GEOID) %>%
  summarise(
    total_filings = sum(filings_2020, na.rm = TRUE),
    median_rent = first(median_rent.y),
    .groups = "drop"
  )

map_sf <- philly_tract_sf %>%
  left_join(map_data, by = "GEOID") %>%
  filter(!is.na(total_filings) & !is.na(median_rent))

# create tertiles for each variable
map_sf <- map_sf %>%
  mutate(
    filings_class = ntile(total_filings, 3),
    rent_class = ntile(median_rent, 3),
    bi_class = paste0(rent_class, "-", filings_class)
  )

# define bivariate color palette (different colors for this one)
bivariate_colors2 <- c(
  "1-1" = "#e8e8e8",  # Low rent, Low evictions
  "2-1" = "#b8d6be",  # Med rent, Low evictions
  "3-1" = "#73ae80",  # High rent, Low evictions
  "1-2" = "#dfb0d6",  # Low rent, Med evictions
  "2-2" = "#a5add3",  # Med rent, Med evictions
  "3-2" = "#5698b9",  # High rent, Med evictions
  "1-3" = "#be64ac",  # Low rent, High evictions
  "2-3" = "#8c62aa",  # Med rent, High evictions
  "3-3" = "#3b4994"   # High rent, High evictions
)

# map
map2 <- ggplot(map_sf) +
  geom_sf(aes(fill = bi_class), color = NA) +
  scale_fill_manual(values = bivariate_colors2, na.value = "grey90", guide = "none") +
  labs(
    title = "Eviction Filings vs Median Rent",
    subtitle = "Darker purple = High evictions AND high rent"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16))

legend_data <- expand.grid(x = 1:3, y = 1:3) %>%
  mutate(bi_class = paste0(x, "-", y))

legend2 <- ggplot(legend_data, aes(x = x, y = y, fill = bi_class)) +
  geom_tile() +
  scale_fill_manual(values = bivariate_colors2, guide = "none") +
  labs(x = "Rent →", y = "Evict. →") +
  theme_void() +
  theme(
    axis.title.x = element_text(size = 9, margin = margin(t = 5)),
    axis.title.y = element_text(size = 9, angle = 90, margin = margin(r = 5))
  ) +
  coord_fixed()

final_plot2 <- ggdraw() +
  draw_plot(map2, 0, 0, 1, 1) +
  draw_plot(legend2, 0.7, 0.05, 0.22, 0.22)

print(final_plot2)
```


# Evication vs Median Household Income
```{r}
# prepare data
map_data <- df_final %>%
  group_by(GEOID) %>%
  summarise(
    total_filings = sum(filings_2020, na.rm = TRUE),
    median_income = first(median_hh_income),
    .groups = "drop"
  )

map_sf <- philly_tract_sf %>%
  left_join(map_data, by = "GEOID") %>%
  filter(!is.na(total_filings) & !is.na(median_income))

# create tertiles for each variable
map_sf <- map_sf %>%
  mutate(
    filings_class = ntile(total_filings, 3),
    income_class = ntile(median_income, 3),
    bi_class = paste0(income_class, "-", filings_class)
  )

# define bivariate color palette (blue-red for income-eviction)
bivariate_colors3 <- c(
  "1-1" = "#e8e8e8",  # Low income, Low evictions
  "2-1" = "#b5c0da",  # Med income, Low evictions
  "3-1" = "#6c83b5",  # High income, Low evictions
  "1-2" = "#f0b8b8",  # Low income, Med evictions
  "2-2" = "#ad9ea5",  # Med income, Med evictions
  "3-2" = "#567994",  # High income, Med evictions
  "1-3" = "#e64b35",  # Low income, High evictions
  "2-3" = "#985356",  # Med income, High evictions
  "3-3" = "#3b4994"   # High income, High evictions
)

# map
map3 <- ggplot(map_sf) +
  geom_sf(aes(fill = bi_class), color = NA) +
  scale_fill_manual(values = bivariate_colors3, na.value = "grey90", guide = "none") +
  labs(
    title = "Eviction Filings vs Median Household Income",
    subtitle = "Orange = High evictions AND low income"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16))

legend_data <- expand.grid(x = 1:3, y = 1:3) %>%
  mutate(bi_class = paste0(x, "-", y))

legend3 <- ggplot(legend_data, aes(x = x, y = y, fill = bi_class)) +
  geom_tile() +
  scale_fill_manual(values = bivariate_colors3, guide = "none") +
  labs(x = "Income →", y = "Evict. →") +
  theme_void() +
  theme(
    axis.title.x = element_text(size = 9, margin = margin(t = 5)),
    axis.title.y = element_text(size = 9, angle = 90, margin = margin(r = 5))
  ) +
  coord_fixed()

final_plot3 <- ggdraw() +
  draw_plot(map3, 0, 0, 1, 1) +
  draw_plot(legend3, 0.7, 0.05, 0.22, 0.22)

print(final_plot3)
```


# Regression
```{r}
#calculate eviction by population
df_final$evibypop=df_final$filings_2020/df_final$total_pop*100
outlier_threshold = df_final %>%
  filter(!is.na(filings_2020)) %>%
  pull(filings_2020) %>%
  quantile(probs = 0.99, na.rm = TRUE)
#filter 2020 April, may and June
df_filtered = df_final %>%
  filter(!year==2020)%>%
  filter(!total_pop==0)%>%
  filter(!is.na(median_rent.y))%>%
  filter(filings_2020 <= outlier_threshold)

set.seed(123)
n <- nrow(df_filtered)

# 70% training, 30% testing
train_indices <- sample(1:n, size = 0.7 * n)
train_data <- df_filtered[train_indices, ]
test_data <- df_filtered[-train_indices, ]
```

```{r}
#run regression based on census characteristic
options(scipen = 999)

model1=lm(evibypop~pct_white+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+aggregate_rooms+pct_family+total_hh+pct_black+pct_asian,data=df_filtered)
summary(model1)

```

```{r}
vif(model1)
```


```{r}
bptest(model1)
```


```{r}
#p-value is way too small, implying more variables may be needed (make sense)

#calculate the RMSE
model_train_1 <- lm(filings_2020~pct_white+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+aggregate_rooms+pct_family+total_hh+pct_black+pct_asian+total_pop,data = train_data)
test_predictions_1 <- predict(model_train_1, newdata = test_data)
rmse_test_1 <- sqrt(mean((test_data$filings_2020 - test_predictions_1)^2,na.rm = TRUE))
rmse_train_1 <- summary(model_train_1)$sigma


plot(model1)
```




```{r}
model2=lm(filings_2020~median_rent.y+foodcpi+electricitycpi+avg_wkly_wage+median_rent.y+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+pct_family+pct_black+pct_asian+comcpi+total_pop,data=df_final)
summary(model2)
```

```{r}
vif(model2)
```

```{r}
bptest(model2)
```

```{r}
#calculate the RMSE
model_train_2 <- lm(filings_2020~median_rent.y+foodcpi+electricitycpi+avg_wkly_wage+median_rent.y+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+pct_family+pct_black+comcpi+total_pop,data = train_data)
test_predictions_2 <- predict(model_train_2, newdata = test_data)
rmse_test_2 <- sqrt(mean((test_data$filings_2020 - test_predictions_2)^2,na.rm = TRUE))
rmse_train_2 <- summary(model_train_2)$sigma


plot(model2)
```

```{r}
model3=lm(filings_2020~median_rent.y+avg_wkly_wage+median_rent.y+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+pct_family+pct_black+electricitycpi+total_pop+MHI_quantile,data=df_final)
summary(model3)
```

```{r}
vif(model3)
```

```{r}
bptest(model3)
```

```{r}
model_train_3 <- glm.nb(filings_2020~median_rent.y+avg_wkly_wage+median_rent.y+median_hh_income+pct_renter+unemployment_rate+median_year_built_renter+pct_family+pct_black+electricitycpi+I(total_pop)+MHI_quantile,data = train_data)
test_predictions_3 <- predict(model_train_3, newdata = test_data)
rmse_test_3 <- sqrt(mean((test_data$filings_2020 - test_predictions_3)^2,na.rm = TRUE))
rmse_train_3 <- summary(model_train_3)$sigma


plot(model3)
```

