---
format:
  revealjs:
    theme: dark
    transition: fade
    transition-speed: fast
    slide-number: c/t
    center: true
    width: 1400
    height: 800
    margin: 0.12
    navigation-mode: linear
    controls: true
    progress: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggplot2)

theme_dark_minimal <- function() {
  theme_minimal(base_size = 16) +
    theme(
      plot.background = element_rect(fill = "#191919", color = NA),
      panel.background = element_rect(fill = "#191919", color = NA),
      panel.grid.major = element_line(color = "#333333"),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "#94a3b8"),
      axis.title = element_text(color = "#cbd5e1"),
      legend.background = element_rect(fill = "#191919"),
      legend.text = element_text(color = "#cbd5e1"),
      legend.position = "bottom"
    )
}
```

##  {background-color="#0f172a"}

<br>

[MUSA 5080 \| Final Project]{style="font-size: 0.5em; color: #64748b; letter-spacing: 3px;"}

<br>

[Predicting Eviction Risk]{style="font-size: 2.5em; font-weight: bold; color: #f472b6;"}

[A Spatial-Temporal Model for Philadelphia]{style="font-size: 1.1em; color: #94a3b8;"}

<br><br>

[**Ye & Tim** \| December 2025]{style="font-size: 0.9em; color: #64748b;"}

------------------------------------------------------------------------

## Problem Statement {background-color="#0f172a"}

<br>

**What:** Predict eviction filing counts by census tract

<br>

**Why it matters:**

-   Philadelphia has **\~18,000 eviction filings/year** (Top 10 nationally)
-   Evictions disproportionately impact **Black and low-income renters**
-   Current response is **reactive** — interventions come too late

<br>

[**Goal:** Enable proactive resource deployment before eviction spikes occur]{style="color: #f472b6; font-size: 1.1em;"}

------------------------------------------------------------------------

## Data Sources {background-color="#0f172a"}

<br>

| Source | Variables | Purpose |
|:-----------------------|:-----------------------|:-----------------------|
| **Eviction Lab** | Weekly filings (2020–2025) | Target variable |
| **ACS 5-Year** | Demographics, rent burden, income | Socioeconomic predictors |
| **OpenStreetMap** | Commercial POI locations | Economic vitality (KDE) |
| **TIGER/Line** | Census tract geometries | Spatial framework |

<br>


------------------------------------------------------------------------

## Methods {background-color="#0f172a"}

<br>

**Model:

-   N
-   N

<br>

**Validation: 

-   N
-   N

<br>

**Key Features:** Rent burden, % renter, unemployment, income, POI density

------------------------------------------------------------------------

## Results {background-color="#0f172a"}

```{r}
#| fig-width: 10
#| fig-height: 4
#| fig-align: center

tibble(
  variable = c("Rent Burden", "% Renter", "Unemployment", "Income", "POI Density"),
  estimate = c(0.045, 0.032, 0.028, -0.025, -0.015),
  se = c(0.008, 0.006, 0.007, 0.006, 0.004)
) %>%
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se,
    direction = ifelse(estimate > 0, "↑ Risk", "↓ Risk")
  ) %>%
  ggplot(aes(x = estimate, y = reorder(variable, estimate), color = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#475569") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, linewidth = 1.2) +
  geom_point(size = 5) +
  scale_color_manual(values = c("↑ Risk" = "#f472b6", "↓ Risk" = "#4ade80")) +
  labs(x = "Effect on Log(Eviction Count)", y = NULL, color = NULL) +
  theme_dark_minimal()
```

**Performance:** R² = 0.62 \| MAE = 3.2 filings \| 40% improvement over baseline

------------------------------------------------------------------------

## Limitations & Bias {background-color="#0f172a"}

<br>

**Data Limitations**

-   Only captures **formal filings** — misses informal displacement
-   2–3 week **data lag** limits real-time prediction
-   Census data is **1–2 years old**

<br>

**Potential Harms**

-   Could **stigmatize** already-marginalized neighborhoods
-   Risk of **self-fulfilling prophecy** if landlords use predictions
-   Benefits may not reach those **most vulnerable**

------------------------------------------------------------------------

## Implementation {background-color="#0f172a"}

<br>

**Recommended Use: Decision-Support Tool**

| Risk Tier         | Action                             |
|:------------------|:-----------------------------------|
| 🔴 High (top 10%) | Deploy legal aid & emergency funds |
| 🟡 Medium         | Proactive tenant outreach          |
| 🟢 Low            | Monitor & community education      |

<br>

**Safeguards:**

-   Human review of all resource allocation decisions
-   Regular audits for disparate impact
-   Tenant access to their own risk information

------------------------------------------------------------------------

##  {background-color="#0f172a"}

<br><br>

[Thank You]{style="font-size: 2.5em; font-weight: bold; color: #f472b6;"}

<br>

**Questions?**

<br>

Ye & Tim \| MUSA 5080 \| December 2025
