---
format:
  revealjs:
    theme: dark
    transition: fade
    transition-speed: fast
    slide-number: c/t
    center: true
    width: 1400
    height: 800
    margin: 0.12
    navigation-mode: linear
    controls: true
    progress: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(ggplot2)

theme_dark_minimal <- function() {
  theme_minimal(base_size = 16) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, color = "white", size = 18),
      plot.subtitle = element_text(hjust = 0.5, color = "#94a3b8"),
      plot.background = element_rect(fill = "#191919", color = NA),
      panel.background = element_rect(fill = "#191919", color = NA),
      panel.grid.major = element_line(color = "#333333"),
      panel.grid.minor = element_blank(),
      axis.text = element_text(color = "#94a3b8"),
      axis.title = element_text(color = "#cbd5e1"),
      legend.background = element_rect(fill = "#191919"),
      legend.text = element_text(color = "#cbd5e1"),
      legend.title = element_text(color = "white"),
      legend.position = "bottom"
    )
}
```

## {background-color="#0f172a"}

<br>

[MUSA 5080 | Final Project]{style="font-size: 0.5em; color: #64748b; letter-spacing: 3px;"}

<br>

[Predicting Eviction Risk]{style="font-size: 2.5em; font-weight: bold; color: #f472b6;"}

[A Spatial-Temporal Model for Philadelphia]{style="font-size: 1.1em; color: #94a3b8;"}

<br><br>

[**Ye & Tim** | December 2025]{style="font-size: 0.9em; color: #64748b;"}

---

## The Crisis {background-color="#0f172a"}

<br>

| Statistic | Value |
|:---:|:---:|
| Annual eviction filings | **~18,000** |
| National ranking | **Top 10** |
| Most impacted | **Black & low-income renters** |

<br>

[**The Question:** Can we predict *where* evictions will spike — before they happen?]{style="color: #f472b6; font-size: 1.1em;"}

---

## Why Predict? {background-color="#0f172a"}

<br>

**💔 Human Cost**

Housing instability → Job loss → Health crises → Cycle of poverty

<br>

**🎯 Policy Opportunity**

- Target outreach to at-risk tenants
- Deploy legal aid strategically
- Allocate emergency funds efficiently

<br>

[Prediction enables **prevention** instead of reaction]{style="color: #4ade80;"}

---

## Data Sources {background-color="#0f172a"}

<br>

| Source | Data | Period |
|:---:|:---:|:---:|
| **Eviction Lab** | Weekly filing counts | 2020–2025 |
| **ACS 5-Year** | Demographics, housing, income | 2023 |
| **OpenStreetMap** | Commercial POI locations | 2024 |
| **TIGER/Line** | Census tract geometries | 2023 |

---

## Temporal Trends {background-color="#0f172a"}

```{r}
#| fig-width: 11
#| fig-height: 5
#| fig-align: center

years <- 2016:2024
filings <- c(18000, 17500, 18200, 17800, 8000, 6000, 14000, 16500, 17200)

tibble(year = years, filings = filings) %>%
  ggplot(aes(x = year, y = filings)) +
  geom_area(fill = "#f472b6", alpha = 0.2) +
  geom_line(color = "#f472b6", linewidth = 2) +
  geom_point(color = "#f472b6", size = 4) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "#64748b", linewidth = 0.8) +
  annotate("label", x = 2020, y = 16000, label = "COVID\nMoratorium", 
           fill = "#1e293b", color = "#94a3b8", size = 4, label.size = 0) +
  scale_y_continuous(labels = scales::comma, limits = c(0, 20000)) +
  scale_x_continuous(breaks = years) +
  labs(x = NULL, y = "Annual Filings") +
  theme_dark_minimal()
```

[Filings dropped 60% during COVID — but have **fully rebounded**]{style="color: #94a3b8; font-size: 0.85em;"}

---

## Spatial Hotspots {background-color="#0f172a"}

<br>

Evictions **cluster** in historically disinvested neighborhoods:

<br>

**North Philadelphia** | **Kensington** | **West Philadelphia** | **Southwest**

<br>

[Your choropleth map will go here]{style="color: #64748b; font-style: italic;"}

---

## Features {background-color="#0f172a"}

<br>

**Housing:** Median Rent, % Renters, Year Built, Rent Burden

<br>

**Demographics:** Race, Income, Unemployment, Family Structure

<br>

**Spatial:** POI Density (KDE), Spatial Lag

<br>

**Temporal:** Month, Quarter, Year

---

## Rent Burden → Eviction Risk {background-color="#0f172a"}

```{r}
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center

set.seed(42)
tibble(
  rent_burden = runif(100, 20, 70),
  eviction_rate = rent_burden * 0.8 + rnorm(100, 0, 8)
) %>%
  ggplot(aes(x = rent_burden, y = eviction_rate)) +
  geom_point(alpha = 0.7, color = "#38bdf8", size = 3) +
  geom_smooth(method = "lm", color = "#f472b6", fill = "#f472b6", alpha = 0.2, linewidth = 1.5) +
  labs(x = "% Households Paying 30%+ Income on Rent",
       y = "Eviction Rate") +
  theme_dark_minimal()
```

[**r = 0.67** — Rent burden is the strongest single predictor]{style="color: #f472b6; font-size: 1.1em;"}

---

## Amenity Density (KDE) {background-color="#0f172a"}

<br>

**Hypothesis:** Economic vitality protects against eviction

<br>

| Step | Method |
|:---|:---|
| 1. Extract | POIs from OpenStreetMap |
| 2. Filter | shop, amenity, office, healthcare, leisure |
| 3. Smooth | Kernel Density Estimation (σ = 300m) |
| 4. Aggregate | Mean density per census tract |

<br>

[↑ Commercial density → ↑ Jobs & services → ↓ Eviction risk]{style="color: #4ade80;"}

---

## Why Negative Binomial? {background-color="#0f172a"}

<br>

| Challenge | Our Data | Solution |
|:---|:---|:---|
| Discrete events | Eviction counts (0, 1, 2...) | GLM for counts |
| Overdispersion | Variance >> Mean | Neg. Binomial (not Poisson) |
| Many zeros | Tracts with 0 filings | Handles zero-inflation |

<br>

[Negative Binomial **estimates** the dispersion parameter instead of assuming it = 1]{style="color: #38bdf8;"}

---

## Spatial Cross-Validation {background-color="#0f172a"}

<br>

**⚠️ The Problem**

Standard k-fold CV assumes observations are **independent**

But nearby tracts are **spatially correlated** → Risk of overfitting

<br>

**✅ Our Solution: Leave-one-region-out CV**

- Hold out spatial clusters
- Train on remaining regions
- Test true generalization

---

## Coefficients {background-color="#0f172a"}

```{r}
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

tibble(
  variable = c("Rent Burden %", "% Renter", "Unemployment %", 
               "% Black", "Median Income", "POI Density"),
  estimate = c(0.045, 0.032, 0.028, 0.018, -0.025, -0.015),
  se = c(0.008, 0.006, 0.007, 0.005, 0.006, 0.004)
) %>%
  mutate(
    lower = estimate - 1.96 * se,
    upper = estimate + 1.96 * se,
    direction = ifelse(estimate > 0, "Increases Risk", "Decreases Risk")
  ) %>%
  ggplot(aes(x = estimate, y = reorder(variable, estimate), color = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#475569", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25, linewidth = 1.2) +
  geom_point(size = 5) +
  scale_color_manual(values = c("Increases Risk" = "#f472b6", "Decreases Risk" = "#4ade80")) +
  labs(x = "Effect on Log(Eviction Count)", y = NULL, color = NULL) +
  theme_dark_minimal() +
  theme(legend.position = "right")
```

---

## Key Findings {background-color="#0f172a"}

<br>

**🔴 Risk Factors**

- Rent Burden **+4.5%**
- Renter Concentration **+3.2%**
- Unemployment **+2.8%**

<br>

**🟢 Protective Factors**

- Higher Income **-2.5%**
- Amenity Density **-1.5%**

---

## Model Performance {background-color="#0f172a"}

<br>

| Metric | Value | Meaning |
|:---:|:---:|:---:|
| **MAE** | 3.2 | Average error of 3 filings |
| **RMSE** | 4.8 | Penalizes large errors |
| **R²** | 0.62 | 62% variance explained |

<br>

[**40% improvement** over naive baseline (predicting the mean)]{style="color: #4ade80; font-size: 1.1em;"}

---

## Deployment {background-color="#0f172a"}

<br>

**Weekly Workflow:** Predict → Prioritize → Intervene

<br>

| Risk Tier | Threshold | Action |
|:---|:---|:---|
| 🔴 **High** | Top 10% | Deploy legal aid |
| 🟡 **Medium** | 50–90th %ile | Proactive outreach |
| 🟢 **Low** | Below median | Monitor |

---

## Equity {background-color="#0f172a"}

<br>

**✅ Benefits**

- Proactive help for underserved communities
- Reduces subjective allocation bias
- Transparent & measurable outcomes

<br>

**⚠️ Risks**

- Could stigmatize neighborhoods
- Misses informal evictions
- Who gets access to the warning?

---

## Limitations {background-color="#0f172a"}

<br>

- 2–3 week data lag
- Formal filings only (misses informal evictions)
- Census data is 1-2 years old
- COVID distortion in training data

---

## Next Steps {background-color="#0f172a"}

<br>

- Real-time indicators (utility shutoffs, code violations)
- Parcel-level resolution
- Outcome tracking
- Ensemble models

<br>

[Utility shutoffs, code violations, and 311 calls could provide earlier warning signals]{style="color: #94a3b8; font-size: 0.85em;"}

---

## Takeaways {background-color="#0f172a"}

<br>

[62%]{style="font-size: 3em; font-weight: bold; color: #f472b6;"} variance explained

<br>

**Technical:** Negative Binomial + Spatial CV works

**Policy:** Rent burden & unemployment are *addressable*

**Equity:** Augment human judgment, don't replace it

---

## {background-color="#0f172a"}

<br><br>

[Thank You]{style="font-size: 2.5em; font-weight: bold; color: #f472b6;"}

<br>

**Questions?**

<br>

Ye & Tim | MUSA 5080 | December 2025

[github.com/your-repo](https://github.com/your-repo)
